---
title: "OSM B: Application"
---


This demo presents four scenarios showcasing how our database of design parameters can be used in power analysis when planning randomized intervention studies on students’ SEL outcomes. We show how to determine:

1) the minimum required sample size (*MRSS*) expressed as the number of students  
   <i class="bi bi-check"></i> illustrated by the example of a single-level individually randomized trial (IRT)

2) the minimum detectable effect size (*MDES*)  
   <i class="bi bi-check"></i> illustrated by the example of a three-level cluster randomized trial (CRT)

3) statistical power (1−β)  
   <i class="bi bi-check"></i> illustrated by the example of a two-level multisite individually randomized trial (MSIRT)

4) the minimum required sample size (*MRSS*) expressed as the number of schools  
   <i class="bi bi-check"></i> illustrated by the example of a three-level multisite cluster randomized trial (MSCRT)


## Prerequisites

### R Setup

We demonstrate power analysis in R using the package PowerUpR [@bulusPowerUpRPowerAnalysis2021]. It implements the power formulas given in Dong and Maynard [-@dongPowerUpToolCalculating2013]. Specifically, we use the functions `mrss.ira()` and `mrss.bcra3r2()` to calculate the *MRSS* (Scenario 1 and 4, respectively), `mdes.cra3()` to calculate the *MDES* (Scenario 2), and `power.bira2()` to calculate 1−β (Scenario 3) for the main treatment effect. Researchers may also use the companion [PowerUpR shiny app](https://powerupr.shinyapps.io/index/) [@atanekaPowerUpRShinyApp2023] or the Excel tool [PowerUp!](https://www.causalevaluation.org/power-analysis.html) [@dongPowerUpToolPower2015]. Note that there are several other appropriate software solutions for planning IRTs and CRTs, such as [Optimal Design](https://wtgrantfoundation.org/optimal-design-with-empirical-information-od) [@spybrookOptimalDesignEmpirical2011].

  


```{r}
#| warning: false
#| output: false

# -- Load packages
# Data wrangling and handling
library(dplyr)
library(stringr)
library(tidyr)
library(purrr)
library(tibble)
# Table processing and plotting
library(knitr)
library(ggplot2)
# Power analysis 
library(PowerUpR)
```


```{r}
#| echo: false

# Setup-Chunk: Automatic reporting functions ---INTERNAL ONLY---

# helper: format esv
format_esv <- function(esv) {
  dplyr::case_when(
    isTRUE(all.equal(esv, .05^2)) ~ "τ² = .05²",
    isTRUE(all.equal(esv, .15^2)) ~ "τ² = .15²",
    isTRUE(all.equal(esv, .35^2)) ~ "τ² = .35²",
    TRUE ~ glue::glue("τ² = {formatC(esv, format = 'f', digits = 4)}")
  )
}

describe_range <- function(data, col, stat = "N", set = NULL, esv_val = NULL) {
  if (!is.null(set)) {
    data <- dplyr::filter(data, Set == set)
  }
  
  if (!is.null(esv_val)) {
    data <- dplyr::filter(data, esv == esv_val)
  }
  
  min_val <- min(data[[col]], na.rm = TRUE)
  max_val <- max(data[[col]], na.rm = TRUE)
  
  min_val <- if (min_val %% 1 == 0) as.character(min_val) else sprintf("%.2f", min_val)
  max_val <- if (max_val %% 1 == 0) as.character(max_val) else sprintf("%.2f", max_val)
  
  suffix <- ""
  if (!is.null(set)) {
    suffix <- paste0(" when controlling for Set ", set)
  }
  
  if (!is.null(esv_val)) {
    suffix <- paste0(" when assuming ", format_esv(esv_val))
  }
  
  if (stat == "1-β") {
  glue::glue("{min_val} ≤ {stat} ≤ {max_val}{suffix}") 
  } else
    glue::glue("{min_val} ≤ *{stat}* ≤ {max_val}{suffix}")
}

describe_min <- function(data, col, stat = "N", ref = TRUE) {
  min_row <- dplyr::slice_min(data, !!rlang::sym(col), n = 1)
  val <- min_row[[col]]
  
  val <- if (val %% 1 == 0) as.character(val) else sprintf("%.2f", val)
  
  suffix <- if (ref) glue::glue(" ({min_row$Method}, {min_row$Set})") else ""
  glue::glue("*{stat}* = {val}{suffix}")
}

describe_max <- function(data, col, stat = "N", ref = TRUE) {
  max_row <- dplyr::slice_max(data, !!rlang::sym(col), n = 1)
  val <- max_row[[col]]
  
  val <- if (val %% 1 == 0) as.character(val) else sprintf("%.2f", val)
  
  suffix <- if (ref) glue::glue(" ({max_row$Method}, {max_row$Set})") else ""
  glue::glue("*{stat}* = {val}{suffix}")
}

pull_inline <- function(data, col, ...) {
  data %>%
    dplyr::filter(...) %>%
    dplyr::pull(.data[[col]])
}
```

```{r}
#| echo: false
urls <- list(
  pe = "https://raw.githubusercontent.com/sophiestallasch/sel-designparams/main/data/dp_pe.pub.rda",
  ma = "https://raw.githubusercontent.com/sophiestallasch/sel-designparams/main/data/dp_ma.pub.rda"
)

if (interactive()) {
  # Local loading for development
  load("../data/dp_pe.pub.rda")
  load("../data/dp_ma.pub.rda")
} else {
  # Remote loading for rendering
  walk(urls, ~ load(url(.x), envir = .GlobalEnv))
}
```

```{r}
#| eval: false
# -- Load design parameter database into global environment
urls <- list(
  # point estimates
  pe = "https://raw.githubusercontent.com/sophiestallasch/sel-designparams/main/data/dp_pe.pub.rda",
  # meta-analytic summaries
  ma = "https://raw.githubusercontent.com/sophiestallasch/sel-designparams/main/data/dp_ma.pub.rda"
)

walk(urls, ~ load(url(.x), envir = .GlobalEnv))
```
If you encounter any issues, download the .rda files for the [point estimates](https://github.com/sophiestallasch/sel-designparams/raw/main/data/dp_pe.pub.rda) 
and [meta-analytic summaries](https://github.com/sophiestallasch/sel-designparams/raw/main/data/dp_ma.pub.rda) and load it manually into R.


### Assumptions

Generally, we assume a balanced design where the respective randomization units are randomly assigned to experimental conditions in equal shares (i.e., 50% of the sample in the treatment group and 50% in the control group; *P* = 0.50). Further, we set the desired power at 80% (1-β = 0.80) and use a two-tailed test with Type I error rate of 5% (α = 0.05) which allows for the detection of potentially unexpected negative intervention effects. Note that these assumptions correspond to the default settings in PowerUpR.


## Scenario 1: Single-level IRT

:::{.scenario-header}
:::{.scenario-subtitle}
A lab-based IRT on SEL outcomes across multiple contexts and domains in elementary school
:::
:::{.scenario-question}
<i class="bi bi-arrow-right-short"></i> How many students are required?
:::
:::

A research team has developed a universal intervention to promote elementary students’ SEL outcomes across multiple contexts and domains. The team plans a single-level lab-based randomized experiment to test whether the new approach is effective under controlled lab conditions across different assessment methods. 

The researchers expect a treatment effect of a size typically reported for previous randomized experiments on SEL outcomes. The meta-analysis by Wilson et al. [-@wilsonEmpiricalBenchmarksEffect2025, Table 2] suggests that the average standardized mean difference across social and behavioral outcomes lies in the range 0.07 ≤ *SMD* ≤ 0.35, with a median of *Mdn*(*SMD*) = 0.22. The team therefore considers a treatment effect of this size on students’ SEL outcomes plausible and meaningful. 

The researchers thus conduct power analysis to determine the required number of students that should be sampled to ensure their IRT can achieve a minimum detectable effect size of *MDES* = 0.22 (with 1-β = 0.80 and α = 0.05 in a two-tailed test).  

```{r}
# set the MDES
es.s1 <- .22
```



#### Baseline Design without Covariates

In a first setup, the research team considers a design without covariates.  

```{r}
# apply PowerUpR::mrss.ira()
mrss.ira(
  es = es.s1,          # Target effect size/MDES
  power = .80,         # Power; default: 80%
  alpha = .05,         # Probability of Type I error; default: 5% 
  two.tailed = TRUE,   # Hypothesis test; default: two-tailed
  p = .50,             # Proportion of students in the treatment group; default: 50%
  n0 = 10,             # Starting value for finding the sample size; default: 10
  tol = .10,           # Tolerance of iterations for finding the sample size; default: 0.1
  g1 = 0,              # Number of covariates; default: 0
  r21 = 0              # Proportion of explained variance by covariates; default: 0
  )
```

```{r}
#| echo: false
#| output: false
d.s1.0 <- mrss.ira(es=es.s1)
```

The *MRSS* equals `r d.s1.0$n`. Therefore, in total, *N* = `r d.s1.0$n` students are required to achieve *MDES* = 0.22 when omitting covariates. Hence, the researchers would have to randomly assign *n* = `r round(d.s1.0$n/2, 0)` students to the treatment group and *n* = `r round(d.s1.0$n/2, 0)` students to the control group.


#### Designs with Covariates

In a second setup, the research team would like to test the impact of different covariate sets (in terms of *R*^2^~Total~) on the required sample size. 

Because the intervention targets SEL outcomes across multiple domains and contexts (as opposed to a specific measure), the researchers rely on the meta-analytic summaries of the design parameters. Specifically, they consult the results of MA-1 with student self-reports, parent ratings, and teacher ratings in elementary school. Note that design parameters are not available for teacher ratings in Model Sets 2-BL and 3-SD+BL.   

Moreover, the team intends to take into account the statistical uncertainty associated with the estimated design parameters. For this purpose, the researchers determine conservative upper/liberal lower bounds for the *MRSS* by drawing on the respective lower/upper bound estimates of the 95% CIs of the meta-analytic averages of *R*^2^~Total~. 




```{r}
# select the relevant design parameters
dp.s1 <- dp_ma %>% 
  filter(Type == "MA-1", 
         Population == "Total",
         Design == "1L-D",
         Edu.level == "Elementary") %>% 
  
  # select relevant columnns
  select(Method, Set, n_Covariates, Parameter, Average, starts_with("CI"))
```

Table 1 shows the design parameters applied in the power analyses:

```{r}
#| code-fold: true
#| label: tbl-dp.s1
#| tbl-cap: Design Parameters for Scenario 1

# inspect selected design parameters
kable(dp.s1 %>% 
        mutate(across(c(Average:CI.ub), ~sprintf("%.2f", .))))
```

Using these estimates, the research team systematically studies how the various covariate sets influence the *MRSS*.

```{r}
#| warning: false 
# prepare database
dp.s1 <- dp.s1 %>% 
  pivot_longer(c(Average:CI.ub), 
               names_to = "statistic", values_to = "r2_total")
```


```{r}
#| warning: false 
#| output: false

# custom function to vectorize PowerUpR::mrss.ira() across varying model sets
vectorize_mrss.ira <- function(g1, r21, ...) {
  parms <- list(es = es.s1, g1 = g1, r21 = r21)
  d_out <- exec(getFromNamespace("mrss.ira", ns = "PowerUpR"), !!!parms)
  d_out$n
}

# apply vectorize_mrss.ira()
d.s1 <- dp.s1 %>% 
  mutate(mrss = map2_dbl(n_Covariates, r2_total, vectorize_mrss.ira))
```

```{r}
#| warning: false
# clean and reshape results 
d.s1 <- d.s1 %>% 
  select(Method, Set, statistic, mrss) %>% 
  mutate(statistic = case_match(statistic,
                                "Average" ~ "MRSS",
                                "CI.lb"   ~ "MRSS conservative",
                                "CI.ub"   ~ "MRSS liberal")) %>%
  pivot_wider(names_from = statistic, values_from = mrss) %>% 
  arrange(Method, Set)
```


Table 2 lists the results; Figure 1 visualizes them:

::: {.grid style="gap: 4rem;"}
::: {.g-col-6}
```{r}
#| code-fold: true
#| label: tbl-d.s1
#| tbl-cap: MRSS for Different Covariate Sets by Assessment Method

# show results
kable(d.s1)
```
:::
::: {.g-col-6}
```{r}
#| code-fold: true
#| label: fig-d.s1
#| fig-cap: MRSS (with conservative/liberal bounds) for Different Covariate Sets by Assessment Method
#| cap-location: top
#| fig-height: 2.9

ggplot(d.s1, aes(x = Set, y = MRSS, fill = Method)) +
  geom_hline(yintercept = d.s1.0$n, linetype = "dashed") + 
  geom_col(position = position_dodge(width = 0.8, preserve = "single"), width = 0.7) +
  geom_errorbar(
    aes(ymin = `MRSS conservative`, ymax = `MRSS liberal`),
    position = position_dodge(width = 0.8, preserve = "single"), 
    width = 0.2
    ) +
  scale_y_continuous(breaks = c(seq(0, 600, 200), d.s1.0$n), limits = c(NA, 700)) + 
  scale_fill_grey(start = 0.3, end = 0.8) +
  labs(y = "MRSS") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(), 
    axis.title.x = element_blank(),
    legend.title = element_blank()
  ) +
  annotate(
    "text", x = Inf, y = d.s1.0$n, label = "Baseline",
    size = 3.5, hjust = 1, vjust = -0.5
  )
```
:::
:::

Figure 1 shows that adding covariates can substantially reduce the *MRSS*. As detailed in Table 2, entering the point estimates of *R*^2^~Total~ into the power formula results in total sample size requirements of `r describe_range(d.s1, "MRSS", "N", "1-SD")`, `r describe_range(d.s1, "MRSS", "N", "2-BL")`, and `r describe_range(d.s1, "MRSS", "N", "3-SD+BL")`, depending on the assessment method. Using the lower bound estimates of the 95% CI of *R*^2^~Total~, and thus, following a conservative approach, the *MRSS* varies between `r describe_min(d.s1, "MRSS conservative", "N")` and `r describe_max(d.s1, "MRSS conservative", "N")`. In contrast, using the respective upper bound estimates to adopt a liberal approach, the *MRSS* varies between `r describe_min(d.s1, "MRSS liberal", "N")` and `r describe_max(d.s1, "MRSS liberal", "N")`.   

To conclude, if covariates cannot be included, a total of *N* = `r d.s1.0$n` elementary students are necessary to achieve *MDES* = `r es.s1`. However, when covariates are included in the design, the required sample size depends on the team's risk preferences. For example, if they opted for a conservative approach by assessing both sociodemographic characteristics and a pretest, the researchers should recruit at least *N* = `r pull_inline(d.s1, "MRSS conservative", Set == "3-SD+BL", Method == "Student")`/`r pull_inline(d.s1, "MRSS conservative", Set == "3-SD+BL", Method == "Parent")`  students to adequately power their IRT to detect *SMD* = `r es.s1` based on student self-reports/parent ratings. 



## Scenario 2: Three-level CRT

:::{.scenario-header}
:::{.scenario-subtitle}
A field three-level CRT on enjoyment of learning among Grade 4 students
:::
:::{.scenario-question}
<i class="bi bi-arrow-right-short"></i> Which MDES is Attainable?
:::
:::

The research team found that their intervention improved 4th graders' self-reported enjoyment of learning with *SMD* = 0.22. The team plans a large-scale trial to evaluate whether the intervention is also effective when implemented by teachers in everyday school routine. To avoid unintentionally exposing the control group to the intervention, the researchers want to carry out a three-level cluster-randomized trial, where whole schools will be randomized to experimental conditions. 

80 schools will participate in the study. From each of these schools, the team intends to sample 2 classrooms, with 20 students per classroom. Given this sample size, the researchers want to determine whether their CRT achieves *MDES* = 0.22 (with 1-β = .80 and α = .05 in a two-tailed test). Since the intervention targets a specific measure, the team enters the point estimates for the design parameters into their power calculations. Again, the researchers allow for the statistical uncertainty associated with the estimated design parameters. They determine conservative upper/liberal lower bound estimates for the *MDES* by drawing on the respective upper/lower bounds of the 95% CIs of the ICCs and the respective lower/upper bounds of the 95% CIs of the *R*^2^ values. 

```{r}
# set the sample size
n.s2 = 20           # Number of students per classroom
J.s2 = 2            # Number of classrooms per school
K.s2 = 80           # Number of schools
```


```{r}
# select the relevant design parameters
dp.s2 <- dp_pe %>% 
  filter(Population == "Total",
         Design == "3L-D",
         Edu.level == "Elementary",
         Grade == 4,
         Method == "Student",
         Measure == "Enjoyment of learning") %>% 
  
  # calculate 95% CIs 
  mutate(CI.lb = Estimate - 1.96*SE,
         CI.ub = Estimate + 1.96*SE) %>% 
  
  # truncate to [0,1], if necessary
  mutate(across(starts_with("CI"), ~ pmax(0, pmin(1, .)))) %>% 
  
  # select relevant columns
  select(Set, n_Covariates, Parameter, Estimate, starts_with("CI"))
```

Table 3 shows the design parameters applied in the power analyses:
```{r}
#| code-fold: true
#| label: tbl-dp.s2
#| tbl-cap: Design Parameters for Scenario 2
# inspect selected design parameters
kable(dp.s2 %>% 
        mutate(across(c(Estimate, starts_with("CI")), ~sprintf("%.2f", .))))
```

#### Baseline Design without Covariates

Again, in a first setup, the research team considers a design without covariates.  

```{r}
# point estimates
mdes.cra3(
  power=.80,         # Power; default: 80%
  alpha=.05,         # Probability of Type I error; default: 5% 
  two.tailed=TRUE,   # Hypothesis test; default: two-tailed
  p=.50,             # Proportion of students in the treatment group; default: 50%
  rho2 = .08,        # Between-classroom differences (ICC_Classroom)
  rho3 = .05,        # Between-school differences (ICC_School)
  g3=0,              # Number of covariates; default: 0
  r21=0,             # Proportion of explained variance by covariates at the student level (R2_Student); default: 0
  r22=0,             # Proportion of explained variance by covariates at the classroom level (R2_Classroom); default: 0
  r23=0,             # Proportion of explained variance by covariates at the school level (R2_School); default: 0
  n = n.s2,          # Number of students per classroom
  J = J.s2,          # Number of classrooms per school
  K = K.s2           # Number of schools
  )

# ICCs: upper bound of the 95% CIs = conservative
mdes.cra3(rho2 = .12, rho3 = .08, n = n.s2, J = J.s2, K = K.s2)

# ICCs: lower bound of the 95% CIs = liberal
mdes.cra3(rho2 = .03, rho3 = .02, n = n.s2, J = J.s2, K = K.s2)
```

```{r}
#| echo: false
#| output: false

d.s2.0 <- mdes.cra3(rho2 = .08, rho3 = .05, n = n.s2, J = J.s2, K = K.s2)   
d.s2.0c <- mdes.cra3(rho2 = .12, rho3 = .08, n = n.s2, J = J.s2, K = K.s2)
d.s2.0l <- mdes.cra3(rho2 = .03, rho3 = .02, n = n.s2, J = J.s2, K = K.s2)
```

When not using covariates, the *MDES* equals `{r} round(d.s2.0$mdes[1], 2)` with the point estimates, `{r} round(d.s2.0c$mdes[1], 2)` with the upper bound estimates of the 95% CIs, and `{r} round(d.s2.0l$mdes[1], 2)` with the lower bound estimates of the 95% CIs of the ICCs. Consequently, with *K* = `{r} K.s2`, the target effect of *SMD* = 0.22 is only detectable under a liberal (but potentially risky) approach to power analysis. 


#### Designs with Covariates

The research team decides to improve design sensitivity by including strong covariates. They therefore proceed to systematically examine how different covariate sets influence the *MDES*.


```{r}
#| warning: false 
# prepare database
dp.s2 <- dp.s2 %>% 
  pivot_longer(c(Estimate:CI.ub), 
               names_to = "statistic", values_to = "r2") %>%
  pivot_wider(names_from = Parameter, values_from = r2) %>% 

# ICC: upper bound of the 95% CI + R2: lower bound of the 95% CI = conservative
# ICC: lower bound of the 95% CI + R2: upper bound of the 95% CI = liberal
  mutate(approach = case_when(
    statistic == "Estimate" ~ "MDES",
    Set == "0" & statistic == "CI.lb" ~ "MDES liberal",
    Set == "0" & statistic == "CI.ub" ~ "MDES conservative",
    Set != "0" & statistic == "CI.lb" ~ "MDES conservative",
    Set != "0" & statistic == "CI.ub" ~ "MDES liberal"))

# extract ICCs
icc <- dp.s2 %>% 
  filter(Set == "0") %>% 
  select(approach, starts_with("ICC"))

dp.s2 <- dp.s2 %>% 
  # fill in ICCs for covariate model sets
  left_join(icc, by = "approach", suffix = c("_", "")) %>% 
  select(Set, n_Covariates, approach, starts_with("ICC"), starts_with("R2"), -ends_with("_")) %>% 
  # remove Model Set 0
  filter(Set != "0")
```


```{r}
#| warning: false 
#| output: false

# custom function to vectorize PowerUpR::mdes.cra3() across varying model sets
vectorize_mdes.cra3 <- function(rho2, rho3, g3, r21, r22, r23, ...) {
  parms <- list(
    rho2 = rho2, rho3 = rho3,           
    g3 = g3, 
    r21 = r21, r22 = r22, r23 = r23,
    n = n.s2, J = J.s2, K = K.s2
  )
  d_out <- exec(getFromNamespace("mdes.cra3", ns = "PowerUpR"), !!!parms)
  d_out$mdes[1]
}

# apply vectorize_mdes.cra3()
d.s2 <- dp.s2 %>% 
  mutate(mdes = pmap_dbl(
    list(
      ICC_Classroom, ICC_School, n_Covariates, R2_Student, R2_Classroom, R2_School
      ), 
    vectorize_mdes.cra3))
```

```{r}
# clean and reshape results
d.s2 <- d.s2 %>% 
  select(-starts_with(c("ICC", "R2"))) %>% 
  pivot_wider(names_from = approach, values_from = mdes)
```



Table 4 lists the results; Figure 2 visualizes them:

::: {.grid style="gap: 4rem;"}
::: {.g-col-6}

```{r}
#| code-fold: true
#| label: tbl-d.s2
#| tbl-cap: MDES for Different Covariate Sets
# show results
kable(d.s2 %>% 
        mutate(across(contains("MDES"), ~sprintf("%.2f", .))))
```
:::
::: {.g-col-6}

```{r}
#| code-fold: true
#| label: fig-d.s2
#| fig-cap: MDES (with conservative/liberal bounds) for Different Covariate Sets
#| cap-location: top
#| fig-height: 1.6
ggplot(d.s2, aes(x = Set, y = `MDES`)) +
  geom_hline(yintercept = es.s1, linetype = "dashed") + 
  geom_segment(aes(xend = Set, y = 0, yend = `MDES`), color = "grey70", linewidth = 1.5) +
  geom_point(aes(y = `MDES`), size = 4, color = "black", fill = "darkgrey", shape = 21) +
  geom_errorbar(
    aes(ymin = `MDES conservative`, ymax = `MDES liberal`),
    width = 0.2
  ) +
  labs(y = "MDES") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
        axis.title.x = element_blank(),
    legend.title = element_blank()
  )  +
  annotate(
    "text", x = Inf, y = es.s1, label = "Target",
    size = 3.5, hjust = 1, vjust = -0.5
  )
```
:::
:::

Figure 2 illustrates that covariates can indeed lower the *MDES*. Table 4 shows that by applying the point estimates, `r describe_range(d.s2, "MDES", "MDES")`, depending on the covariate set. Entering the upper bound estimates of the 95% CIs of the ICCs plus the lower bound estimates of the 95% CIs of the *R*^2^s under a conservative approach, the *MDES* varies between `r describe_min(d.s2, "MDES conservative", "MDES", ref=FALSE)` and `r describe_max(d.s2, "MDES conservative", "MDES", ref=FALSE)`. In contrast, using the respective the lower bound estimates of the 95% CIs of the ICCs in combination with the upper bound estimates of the 95% CIs of *R*^2^s under a liberal approach, the *MDES* varies between `r describe_min(d.s2, "MDES liberal", "MDES", ref=FALSE)` and `r describe_max(d.s2, "MDES liberal", "MDES", ref=FALSE)`. 

In summary, with a fixed maximum of *K* = `r K.s2` schools in total, the research team should at least consider adding a baseline measure, hence, to pretest students' enjoyment of learning, given statistical uncertainty. If doing so, the researchers can be confident that the sample size of their CRT will be large enough to detect *SMD* = `r es.s1` based on student self-reports.



## Scenario 3: Two-level MSIRT

:::{.scenario-header}
:::{.scenario-subtitle}
A field two-level MSIRT on self-esteem among Grade 13 students
:::
:::{.scenario-question}
<i class="bi bi-arrow-right-short"></i> Has the design adequate statistical power?
:::
:::

A research team aims to investigate whether an online intervention can enhance self-esteem among adolescents in Grade 13 of upper secondary school. Because students at this education level typically attend subject-specific courses with varying compositions rather than intact classrooms, the team designs a two-level MSIRT in which students are randomly assigned to experimental conditions within schools forming the sites.

To define a reasonable target effect size, the researchers consult the meta-analysis by Cipriano et al. [-@ciprianoStateEvidenceSocial2023, Table 3], which reports an average standardized mean difference of *SMD* = 0.21 for interventions targeting SEL attitudes and beliefs. Accordingly, the team specifies a minimum detectable effect size of *MDES* = 0.21.

Due to financial constraints, the study needs to be designed as parsimoniously as possible with respect to sample size. The researchers plan to recruit approximately 50 students per school. In contrast, sampling schools is substantially more costly, and the total number of schools is therefore limited to a maximum of 40. To increase design sensitivity, the team decides to collect a pretest measure of self-esteem at the end of Grade 12 and include it as a covariate in the analysis.

For the power calculations, the researchers rely on the point estimates of the design parameters for self-esteem in Grade 13. Beyond these standard design parameters, power calculations for multisite experiments require additional assumptions about the degree of variation in treatment effects across sites. Accordingly, the team also has to specify values for the treatment effect heterogeneity across schools and the proportion of this between-school variance in treatment effects that can be explained by the school-level pretest. Based on Weiss et al. [-@weissHowMuchEffects2017], the standard deviation of standardized treatment effects across sites typically ranges from 0 ≤ τ ≤ .35. To assess the robustness of the study design, the research team therefore considers scenarios with small (τ² = .05²), moderate (τ² = .15²), and large (τ² = .35²) treatment effect variability across schools. Taking a conservative approach, the researchers assume that the pretest explains only a small fraction of the between-school variance in treatment effects—specifically, one tenth of the proportion of variance explained in school-level differences in self-esteem.


```{r}
# set the MDES
es.s3 <- .21
```

```{r}
# set the sample size
n.s3 <- 50             # Number of students per school
K.s3 <- seq(5, 40, 1)  # Realistically feasible range for the number of schools
```

```{r}
# set number of covariates: 1 pretest (Set 2-BL)
ncov.s3 <- 1
```

```{r}
# set the effect size variability: small, moderate, and large 
esv.s3 <- c(.05, .15, .35)^2
```

```{r}
# set the factor for the proportion of variance in treatment effects explained by a pretest at the school level: 10% of R2_School 
fct_R2T.s3 <- .10
```


```{r}
#| eval: false
# load point estimates
load(url("https://raw.githubusercontent.com/sophiestallasch/sel-designparams/main/data/dp_pe.pub.rda"))
```
If you encounter any issues, [click here to download the .rda file](https://github.com/sophiestallasch/sel-designparams/raw/main/data/dp_pe.pub.rda) and load it manually into R.


```{r}
# select the relevant design parameters
dp.s3 <- dp_pe %>% 
  filter(Population == "Total",
         Design == "2L-D",
         Edu.level == "Upper secondary",
         Grade == 13,
         Method == "Student",
         Measure == "Self-esteem", 
         Set == "0" | Set == "2-BL") %>% 
  
  # select relevant columns
  select(Set, n_Covariates, Parameter, Estimate)
```


Table 5 shows the design parameters applied in the power analyses:

```{r}
#| code-fold: true
#| label: tbl-dp.s3
#| tbl-cap: Design Parameters for Scenario 3

# inspect selected design parameters
kable(dp.s3 %>% 
        mutate(across(Estimate, ~sprintf("%.2f", .))))
```

Using these estimates, the research team systematically conducts power analyses to identify the most parsimonious, while still adequately powered design under varying assumptions about τ².


```{r}
#| warning: false 
# extract design parameters
dp.s3 <- dp.s3 %>% 
  select(Parameter, Estimate) %>% 
  deframe

# create tibble that combines esv3 and K
parms.s3 <- crossing(esv = esv.s3, K = K.s3)
```


```{r}
#| warning: false 
#| output: false

# custom function to vectorize PowerUpR::power.bira2() across different assumptions on the effect size variability and varying sample sizes
vectorize_power.bira2 <- function(esv2, K) {
  parms <- list(
    es = es.s3,
    rho2 = dp.s3["ICC_School"],
    esv2 = esv2,                        # Effect size variability at the school level
  # omega2 = esv2/dp.s3["ICC_School"],  # Alternatively, treatment effect heterogeneity: omega2 = esv2/rho2. In this case, ´esv2´ is ignored.
    g2 = ncov.s3, 
    r21 = dp.s3["R2_Student"], 
    r2t2 = dp.s3["R2_School"] * fct_R2T.s3, # Proportion of explained treatment effect variance by covariates at the school level: 1/10 of R2_School; default: 0
    n = n.s3, 
    J = K
  )
  d_out <- exec(getFromNamespace("power.bira2", ns = "PowerUpR"), !!!parms)
  d_out$power
}

# apply vectorize_power.bira2()
d.s3 <- parms.s3 %>% 
  mutate(power = map2_dbl(esv, K, vectorize_power.bira2)) %>% 
  # categorize esv
  mutate(
  esv_label = factor(esv,
    labels = map_chr(as.numeric(levels(factor(esv))), format_esv)
    )
  )
```


Table 6 lists the results; Figure 3 visualizes them:

::: {.grid style="gap: 4rem;"}
::: {.g-col-6}
```{r}
#| code-fold: true
#| label: tbl-d.s3
#| tbl-cap: Power 1−β Across Varying Numbers of Schools and Effect Size Variability 

# show results
d.s3 %>% 
  filter(K %in% c(5, 6, 7, 38, 39, 40)) %>%
  select(-esv) %>% 
  pivot_wider(names_from = esv_label, values_from = power) %>%
  mutate(K = as.character(K), 
         across(starts_with("τ²"), ~sprintf("%.4f", .))) %>%
  {bind_rows(., slice(., 1) %>% mutate(across(everything(), ~ "⋮")))} %>% 
  mutate(K = factor(K, levels = c("5", "6", "7", "⋮", "38", "39", "40"))) %>% 
  arrange(K) %>% 
  kable
```
:::
::: {.g-col-6}
```{r}
#| code-fold: true
#| label: fig-d.s3
#| fig-cap: Power Curves Across Varying Numbers of Schools and Effect Size Variability
#| cap-location: top
#| fig-height: 2.9

k80 <- d.s3 %>%
  group_by(esv_label) %>%
  filter(power >= .80) %>%
  slice_min(K, n = 1) %>%
  ungroup()

ggplot(d.s3, 
       aes(x = K, y = power, color = esv_label)) +
  geom_hline(yintercept = .80, linetype = "dashed") + 
  geom_line(linewidth = 1) +
  geom_segment(data = k80, 
               aes(x = K, xend = K, y = -.025, yend = power, color = esv_label),
               linewidth = 0.75, linetype = "dashed") +
  geom_text(data = k80,
            aes(x = K, y = -.025, label = K),
            color = "#111111", vjust = 1.2, size = 3) +
  scale_y_continuous(breaks = seq(0, 1, .20), limits = c(-.025, 1)) +
  scale_color_grey(start = 0.3, end = 0.8) +
  labs(y = "Power 1−β") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.x = element_blank(),
    legend.title = element_blank()
  ) +
  annotate(
    "text", x = Inf, y = .80, label = "Target",
    size = 3.5, hjust = 1, vjust = -0.5
  )
```
:::
:::


As shown in Table 6, statistical power varies between `r describe_range(d.s3, "power", "1-β", esv = .05^2)`, between `r describe_range(d.s3, "power", "1-β", esv = .15^2)`, and between `r describe_range(d.s3, "power", "1-β", esv = .35^2)`. 

Figure 3 reveals that statistical power increases monotonically with the number of sampled schools across all assumptions about effect size variability. Under the small effect size variability assumption (`r k80$esv_label[1]`), the target power of 80% is achieved with *K* = `r k80$K[1]` schools. Assuming moderate effect size variability (`r k80$esv_label[2]`), at least *K* = `r k80$K[2]` schools are required to reach adequate power. Even under the most conservative assumption of large effect size variability (`r k80$esv_label[3]`), the 80% power threshold is exceeded with *K* = `r k80$K[3]` schools.

Overall, the research team learns that the maximally feasible sample size of *K* = 40 schools is sufficient to ensure adequate statistical power even under pessimistic assumptions regarding treatment effect heterogeneity.



## Scenario 4: Three-level MSCRT

:::{.scenario-header}
:::{.scenario-subtitle}
A field three-level MSCRT on problem behavior in non-academic lower secondary schools
:::
:::{.scenario-question}
<i class="bi bi-arrow-right-short"></i> How many schools are required?
:::
:::

A research team is commissioned by policymakers to evaluate the effectiveness of a newly developed program aimed at reducing problem behaviors—such as violence and bullying—among students in Grades 5 to 10 attending the non-academic track of lower secondary education. Because the intervention targets classroom-level peer processes, the team plans a three-level MSCRT randomizing intact classrooms within schools.

The policymakers indicate that they would support a regular implementation of the program only if its effects are at least comparable to those typically achieved by existing similar interventions. Accordingly, the research team defines a minimum relevant effect size of *MRES* = 0.16, corresponding to the average standardized mean difference reported for interventions targeting externalizing behaviors in the meta-analysis by Cipriano et al. [-@ciprianoStateEvidenceSocial2023, Table 3].

In line with common organizational structures in lower secondary schools, the team plans to sample at least 4 classrooms per school, with 20 students per classroom. To improve statistical precision, the analyses will adjust for students’ sociodemographic characteristics.

To inform the power analyses, the researchers draw on the meta-analytically summarized design parameters for other-oriented SEL domains in a general context from MA-3. To explicitly account for uncertainty in these design parameters, they additionally incorporate the 95% CIs and 95% PIs into their calculations. Specifically, they consider a pessimistic planning scenario by combining the respective upper bounds for ICCs with the lower bounds of the *R*^2^ values. Moreover, the team assumes a moderate degree of cross-school treatment effect heterogeneity, setting the effect size variability to τ² = 0.15² [@weissHowMuchEffects2017]. Further, they assume that school-level sociodemographics can explain a substantial share of this heterogeneity [see @schochetUnderstandingVariationTreatment2014; @weissConceptualFrameworkStudying2014], amounting to approximately half of the variance in between-school differences in other-oriented outcome measures accounted for by these covariates.

In applied school settings, perfectly balanced treatment allocation at the classroom level is often difficult to achieve. Practical constraints, differential opportunities for program implementation, and ethical considerations may result in unequal proportions of classrooms assigned to the treatment and control conditions. To reflect this realistic planning uncertainty, the research team conducts sensitivity analyses across varying treatment allocation proportions in the range of 0.50 ≤ *P* ≤ 0.90. The goal of these power analyses is to determine the minimum required number of schools needed to ensure sufficient statistical power (1-β = 0.80) to detect the policy-relevant MRES of 0.16 (with α = 0.05 in a two-tailed test), even under substantial treatment allocation imbalance.



```{r}
# set the MRES
es.s4 <- .16
```

```{r}
# set the sample size
n.s4 <- 20           # Number of students per classroom
J.s4 <- 4            # Number of classrooms per school
```

```{r}
# set the effect size variability 
esv.s4 <- .15^2      # moderate esv
```

```{r}
# set the factor for the proportion of variance in treatment effects explained by sociodemographics at the school level: 50% of R2_School 
fct_R2T.s4 <- .50
```

```{r}
# set the range of allocation proportions 
P.s4 <- seq(.50, .90, .01) # balanced design: P = 50%; unbalanced designs: P > 50% (treatment or control group overrepresented)
# Note that in this two-arm design, allocation proportions P and 1-P yield identical results due to the (assumed) symmetry of the treatment and control group 
```


```{r}
# select the relevant design parameters
dp.s4 <- dp_ma %>% 
  filter(Type == "MA-3", 
         Population == "Non-academic",
         Design == "3L-D",
         Edu.level == "Lower secondary", 
         Method == "Student", 
         Domain == "Other", 
         Context == "General", 
         Set == "0" | Set == "1-SD") %>% 
  
  # select relevant columnns
  select(Set, n_Covariates, Parameter, Average, starts_with(c("CI", "PI"))) %>% 
  # keep upper bounds of the 95% CI/PI for ICCs + lower bounds of the 95% CI/PI for R2 = most pessimistic case
  mutate(across(contains("ub"), ~ if_else(grepl("^ICC_", Parameter), ., NA_real_)),
         across(contains("lb"), ~ if_else(grepl("^R2_", Parameter), ., NA_real_))) %>% 
  arrange(Set)
```

Table 7 shows the design parameters applied in the power analyses:

```{r}
#| code-fold: true
#| label: tbl-dp.s4
#| tbl-cap: Design Parameters for Scenario 4

# inspect selected design parameters
kable(dp.s4 %>% 
        mutate(across(where(is.double), ~ if_else(is.na(.), "", sprintf("%.2f", .)))))
```

Entering these estimates, the research team conducts a series of power analyses to determine the *MRSS* and to test how sensitive their design reacts towards potential treatment allocation imbalance.

```{r}
#| warning: false 
# prepare database
dp.s4 <- dp.s4 %>% 
  pivot_longer(c(Average:PI.ub), names_to = "statistic", values_drop_na = TRUE) %>%
  pivot_wider(names_from = Parameter, values_from = value) %>% 
  
  # clean `statistic`
  mutate(statistic = str_replace(statistic, "^(CI|PI)\\..*", "95% \\1 (conservative)")) %>% 
  
  # fill in ICCs for covariate model sets
  fill(ICC_Classroom, ICC_School, .by = statistic, .direction = "down") %>% 
  # remove Model Set 0
  filter(Set != "0") %>% 

  # select relevant columns
  select(-Set) 

# add range of allocation proportions
dp.s4 <- dp.s4 %>% 
  crossing(P.s4)
```


```{r}
#| warning: false
#| output: false
# custom function to vectorize PowerUpR::mrss.bcra3r2() across varying treatment allocation proportions
vectorize_mrss.bcra3r2 <- function(rho2, rho3, P, g3, r21, r22, r23,...) {
  parms <- list(
    es = es.s4,
    rho2 = rho2, 
    rho3 = rho3,
    esv3 = esv.s4,
    p = P,
    g3 = g3,
    r21 = r21, 
    r22 = r22, 
    r2t3 = r23 * fct_R2T.s4,
    n = n.s4, 
    J = J.s4
  )
  d_out <- exec(getFromNamespace("mrss.bcra3r2", ns = "PowerUpR"), !!!parms)
  d_out$K
}

# apply vectorize_mrss.bcra3r2()
d.s4 <- dp.s4 %>% 
  mutate(mrss = pmap_dbl(
    list(
      ICC_Classroom, ICC_School, P.s4, n_Covariates, R2_Student, R2_Classroom, R2_School
      ), 
    vectorize_mrss.bcra3r2))
```

```{r}
# clean and reshape results
d.s4 <- d.s4 %>% 
  select(-starts_with(c("n_", "ICC", "R2"))) %>% 
  pivot_wider(names_from = statistic, values_from = mrss) %>% 
  select(P = P.s4, Average, starts_with("95%"))
```

Table 8 lists the results; Figure 4 visualizes them:

::: {.grid style="gap: 4rem;"}
::: {.g-col-6}
```{r}
#| code-fold: true
#| label: tbl-d.s4
#| tbl-cap: MRSS for Average and Conservative Design Parameters by Treatment Allocation Proportion

# show results
d.s4[seq(1, nrow(d.s4), 10), ] %>% 
  kable
```
:::
::: {.g-col-6}
```{r}
#| code-fold: true
#| label: fig-d.s4
#| fig-cap: MRSS (with 95% CI and PI) for Varying Treatment Allocation Proportions
#| cap-location: top
#| fig-height: 2.9
d.s4 %>% 
ggplot(aes(x = P.s4)) +
  geom_ribbon(aes(ymin = Average, ymax = `95% CI (conservative)`, fill = "95% CI"), alpha = 0.5) + 
  geom_ribbon(aes(ymin = Average, ymax = `95% PI (conservative)`, fill = "95% PI"), alpha = 0.5) + 
  geom_line(aes(y = Average, color = "Average"), linewidth = 0.5) + 
  geom_point(aes(y = Average, color = "Average")) +
  
  scale_y_continuous(breaks = seq(20, 90, 10), limits = c(20, 90)) +
  scale_color_manual(values = c("Average" = "#4D4D4D")) +
  scale_fill_manual(values = c("95% CI" = "#9C9C9C", "95% PI" = "#CCCCCC"),
                    breaks = c("95% PI", "95% CI")) +
  guides(fill = guide_legend(order = 1), color = guide_legend(order = 2)) +
  labs(y = "MRSS") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.x = element_blank(),
    legend.title = element_blank(), 
    legend.spacing.y = unit(-0.2, "cm")
  )
```
:::
:::


As shown in Table 8, across allocation proportions, the *MRSS* ranges between `r describe_range(d.s4, "Average", stat = "K")` schools when using the meta-analytic average design parameters, between `r describe_range(d.s4, "95% CI (conservative)", stat = "K")` schools when relying on conservative 95% CI bounds, and between `r describe_range(d.s4, "95% PI (conservative)", stat = "K")` schools when adopting the most pessimistic assumptions based on 95% PI bounds. Under mild imbalance (*P* = .60), the required number of schools equals *K* = `r pull_inline(d.s4, "Average", P == .60)` when using the meta-analytic averages. This requirement increases to *K* = `r pull_inline(d.s4, "95% CI (conservative)", P == .60)` schools when using 95% CI bounds and to *K* = `r pull_inline(d.s4, "95% PI (conservative)", P == .60)` schools when relying on 95% PI bounds. Under strong imbalance (*P* = .80), the *MRSS* correspondingly rises to *K* = `r pull_inline(d.s4, "Average", P == .80)`/`r pull_inline(d.s4, "95% CI (conservative)", P == .80)`/`r pull_inline(d.s4, "95% PI (conservative)", P == .80)` schools.

Figure 4 illustrates that the *MRSS* increases monotonically with growing treatment allocation imbalance in all considered cases. As the allocation proportion departs from a balanced design, increasingly more schools are required to compensate for the resulting loss in statistical efficiency. This increase is modest when relying on the meta-analytic averages, but becomes substantially more pronounced under conservative 95% CI- and 95% PI-based assumptions.

To sum up, the results indicate that treatment allocation imbalance can considerably inflate the *MRSS* in the three-level MSCRT on problem behaviors. Planning under realistic degrees of uncertainty therefore necessitates a noticeably larger school sample to ensure that the design is sensitive enough to detect *MRES* = .16, particularly when strong imbalance is anticipated.

